---
title: "lab 13"
author: "Adelaide Cutler"
format:
  html:
    toc: true
    code-fold: true
    theme: cerulean
    embed-resources: true
    backgroundcolor: "pink"
execute:
  echo: true
  freeze: auto
---
```{r}
.libPaths(c("/home/ascutler_umass_edu/R/x86_64-pc-linux-gnu-library/4.5", "/usr/local/lib/R/site-library", "/usr/local/lib/R/library"))
```

```{r}
library(tidyverse)
library(janitor)
library(plotly)
library(DT)
library(lubridate)
```
# Step 1: Overall Data
```{r}
NEON_soilMAGs_soilChem <- read.csv("NEON_soilMAGs_soilChem.csv")
head(NEON_soilMAGs_soilChem)
```
# Step 2: Taxa distributions 
For lab today I will be going over Bio-geography & Spatial Patterns within the above data set.
These patterns look at the location in relationships with patterns. The below is mapped taxa distributions, which is the mapped abundance of different taxa in the data at certain locations.
```{r}
ggplot(NEON_soilMAGs_soilChem, aes(x = decimalLongitude, y = decimalLatitude, color = phylum)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Taxa distribution of MAGs",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```
Above is a spatial map of taxa by longitude and latitue and colored by type.

```{r}
taxa_by_site <- NEON_soilMAGs_soilChem %>%
  group_by(site_ID, phylum, decimalLatitude, decimalLongitude) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(site_ID) %>%
  mutate(rel_abundance = count / sum(count))

ggplot(taxa_by_site, aes(x = decimalLongitude, y = decimalLatitude, size = rel_abundance, color = phylum)) +
  geom_point(alpha = 0.7) +
  labs(title = "Relative abundance of taxa by site") +
  theme_minimal()

```
This map is of abundance per site of each phylum.

```{r}
library(maps)

world <- map_data("world")

ggplot() +
  geom_polygon(data = world, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray70") +
  geom_point(data = NEON_soilMAGs_soilChem, aes(x = decimalLongitude, y = decimalLatitude, color = phylum),
             size = 3, alpha = 0.7) +
  coord_fixed(1.3) +
  labs(title = "Taxa distribution of MAGs") +
  theme_minimal()

```
This map shows the abundance on a real map of the world so theres a better idea of where everything is located for the reader.

# Step 3: Spatial autocorrelation
This will look to see if geography is associated to phylum communities.
```{r}
coords <- NEON_soilMAGs_soilChem %>%
  select(decimalLongitude, decimalLatitude) %>%
  filter(!is.na(decimalLongitude), !is.na(decimalLatitude))
mags_clean <- NEON_soilMAGs_soilChem %>%
  filter(!is.na(decimalLongitude), !is.na(decimalLatitude))

coords <- mags_clean %>% select(decimalLongitude, decimalLatitude)

library(spdep)

neighbors <- knearneigh(coords, k = 4)
nb <- knn2nb(neighbors)
listw <- nb2listw(nb)

```
The code above will define what counts as neighbors spatialy.
```{r}
mags_clean <- NEON_soilMAGs_soilChem %>% filter(!is.na(soilInWaterpH))
moran.test(mags_clean$soilInWaterpH, listw)
```
This result is a strong correlation meaning similar pH soil values tend to be near each other. This could mean the clustering shows there is an environmental reason behind the clusterings.

# Step 4: Distance-decay analysis
This will look at how the similarity of communities decreases as distance increases.
```{r}
comm_matrix <- NEON_soilMAGs_soilChem %>%
  group_by(site_ID, phylum) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = phylum, values_from = count, values_fill = 0)

# Extract coordinates
coords <- NEON_soilMAGs_soilChem %>%
  distinct(site_ID, decimalLatitude, decimalLongitude) %>%
  arrange(site_ID)

```
```{r}
library(vegan)

# Remove siteID column to get numeric matrix
dist_comm <- vegdist(comm_matrix[,-1], method = "bray")
```
The above code will Bray–Curtis dissimilarity, which is a statistical measure of dissimilarity.
```{r}
dist_geo <- dist(coords[,c("decimalLongitude","decimalLatitude")])
```
The above shows the distance between sites.
```{r}
comm_matrix <- NEON_soilMAGs_soilChem %>%
  group_by(site_ID, phylum) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = phylum, values_from = count, values_fill = 0)
coords <- NEON_soilMAGs_soilChem %>%
  distinct(site_ID, decimalLatitude, decimalLongitude) %>%
  arrange(site_ID)
site_data <- comm_matrix %>%
  inner_join(coords, by = "site_ID")
nrow(comm_matrix)
nrow(coords)
nrow(site_data)

site_data_clean <- na.omit(site_data)
colSums(is.na(site_data_clean))
```
```{r}
library(vegan)
comm_matrix <- site_data_clean %>%
  select(-site_ID, -decimalLatitude, -decimalLongitude)
dist_comm <- vegdist(comm_matrix, method = "bray")
coords <- site_data_clean %>% select(decimalLongitude, decimalLatitude)
dist_geo <- dist(coords)
mantel(dist_comm, dist_geo, method = "pearson")
```
The above results for mantel show a moderately positive correlation between dissimilarity and distance with 0.4057 = r. The significance of p = 0.001 shows a high significance meaning the null hypothesis can be rejected and there is a correlation. Overall these results prove there is a relationship between geographical distance and dissimilarity.












